---
title: "The Four Causes of Adaptation"
subtitle: "How do students construct biological explanations?"
author: "Lucy E Delaney"
institute: "University of Illinois at Chicago"
date: "`r Sys.Date()`"
event: "MEEC, 2021"
output:
  xaringan::moon_reader:
    seal: false
    css:
      - css/manual-fonts.css
      - css/mytheme.css
    includes:
      after_body: css/insert-logo.html
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: right, middle, bg_cherry
# The Four Causes of `r htmltools::withTags(br())` Adaptation

`r htmltools::withTags(br())`

.pull-right[
.border[
We do not have knowledge of a thing until we have grasped its `r htmltools::withTags(b("why"))`, that is to say, its `r htmltools::withTags(b("cause"))`.]

Aristotle, `r htmltools::withTags(br())` _Physics: Book II_]

.pull-bottom[`r rmarkdown::metadata$author` · `r rmarkdown::metadata$event`]

---
class: left, middle, bg_munich

.pull-left[## Lucy Delaney 

`r htmltools::withTags(img(style = "border-radius: 50%;", src = "https://ledelaney.org/talks/sabertalk/css/images/Delaney-Headshot-Original.jpg", width = "170px"))`

[`r fontawesome::fa("envelope")` ldelan5@uic.edu](mailto:ldelan5@uic.edu)  
[`r fontawesome::fa("twitter")` @emergentlyLucy](https://twitter.com/emergentlyLucy)  
[`r fontawesome::fa("github")` @ledelaney](https:://github.com/ledelaney)]

--

.pull-right[## About me

`r emo::ji("mortar_board")` .medium-font[PhD candidate in Evolutionary Biology]
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r emo::ji("hibiscus")` .medium-font[Studying plant sex macroevolution]
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r emo::ji("hatching_chick")` .medium-font[Interested in better biology education]]

--

.pull-bottom[
Link to all slide materials: https://github.com/ledelaney/03-21-MEEC]

---
class: left, bg_totoro


--


## Biology education has a problem
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`

.footnote[
Momsen _et al._, 2010]

--

`r htmltools::HTML('<i class="far fa-tired"></i>')`
`r htmltools::HTML("&nbsp;")`
Biology is `r htmltools::HTML("&ldquo;")`just memorization!`r htmltools::HTML("&rdquo;")`

`r htmltools::withTags(br())`

.footnote[
Momsen _et al._, 2010]

--

`r htmltools::HTML('<i class="far fa-grin-stars"></i>')`
`r htmltools::HTML("&nbsp;")`
Formal instruction is not always leading to robust conceptual understanding

`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`

.footnote[
Momsen _et al._, 2010; Coley `r htmltools::withTags("&amp;")` Tanner, 2015]


---
class: middle, center, bg_gorg

--

.mezzo-forte-font[
M `r htmltools::HTML("&nbsp;")` EECN `r htmltools::HTML("&nbsp;")` AS `r htmltools::HTML("&nbsp;")` ADN `r htmltools::HTML("&nbsp;")` ANC `r htmltools::HTML("&nbsp;")` AA `r htmltools::HTML("&nbsp;")` AC `r htmltools::HTML("&nbsp;")` LUM `r htmltools::HTML("&nbsp;")` K]

`r htmltools::withTags(br())`
`r htmltools::withTags(br())`

Countdown:

`r htmltools::HTML('<span id="demo1">10 </span> Seconds')`
`r htmltools::HTML('<button type="button" class="btn btn-outline-info" onclick="myTimerObj1.start()">Start counter</button>')`

.footnote[
Heath `r htmltools::withTags("&amp;")` Heath, 2007]

---
class: middle, center, bg_gorg

.footnote[
Heath `r htmltools::withTags("&amp;")` Heath, 2007]

--

.mezzo-forte-font[
M `r htmltools::HTML("&nbsp;")` EECN `r htmltools::HTML("&nbsp;")` AS `r htmltools::HTML("&nbsp;")` ADN `r htmltools::HTML("&nbsp;")` ANC `r htmltools::HTML("&nbsp;")` AA `r htmltools::HTML("&nbsp;")` AC `r htmltools::HTML("&nbsp;")` LUM `r htmltools::HTML("&nbsp;")` K]

.footnote[
Heath `r htmltools::withTags("&amp;")` Heath, 2007]

---
class: middle, center, bg_gorg

.footnote[
Heath `r htmltools::withTags("&amp;")` Heath, 2007]

--
.mezzo-forte-font[
MEEC `r htmltools::HTML("&nbsp;")` NASA `r htmltools::HTML("&nbsp;")` DNA `r htmltools::HTML("&nbsp;")` NCAA `r htmltools::HTML("&nbsp;")` ACLU `r htmltools::HTML("&nbsp;")` MLK]

`r htmltools::withTags(br())`
`r htmltools::withTags(br())`

Countdown:

`r htmltools::HTML('<span id="demo2">10 </span> Seconds')`
`r htmltools::HTML('<button type="button" class="btn btn-outline-info" onclick="myTimerObj2.start()">Start counter</button>')`


.footnote[
Heath `r htmltools::withTags("&amp;")` Heath, 2007]

---
class: middle, center, bg_gorg

.footnote[
Heath `r htmltools::withTags("&amp;")` Heath, 2007]

--

.mezzo-forte-font[
MEEC `r htmltools::HTML("&nbsp;")` NASA `r htmltools::HTML("&nbsp;")` DNA `r htmltools::HTML("&nbsp;")` NCAA `r htmltools::HTML("&nbsp;")` ACLU `r htmltools::HTML("&nbsp;")` MLK]

.footnote[
Heath `r htmltools::withTags("&amp;")` Heath, 2007]

--

.pull-bottom[
.large-font[
.blue[Remembering concepts is far easier than remembering raw data!]]]

.footnote[
Heath `r htmltools::withTags("&amp;")` Heath, 2007]

---
class: middle, center, bg_sakcherry

`r htmltools::withTags(br())`

## Evolutionary theory is the concept behind the raw data

.footnote[
Dobzhansky, 1973]

--

.large-font[
🦠 🌾 🐡 🪲 🦧]


.footnote[
Dobzhansky, 1973]
---
class: middle, bg_mountain1

---
class: middle, bg_mountain2

---
class: middle, bg_mountain3

---
class: middle, bg_mountain4

---
class: middle, bg_mountain5

.footnote[
Göransson _et al._, 2020]

---
class: middle, bg_mountain6

.footnote[
Göransson _et al._, 2020]

---
class: middle, bg_mountain7

.footnote[
Göransson _et al._, 2020; Southerland _et al._, 2001]

---
class: middle, bg_mountain8

.footnote[
Göransson _et al._, 2020; Southerland _et al._, 2001; Mayr, 1961]

---
class: middle, left, bg_peacock

--
`r htmltools::HTML('<i class="fas fa-tools"></i>')` 
`r htmltools::HTML("&nbsp;")`
How do students construct biological explanations about complex phenomena?  


--
`r htmltools::HTML('<i class="fas fa-chart-line"></i>')`
`r htmltools::HTML("&nbsp;")`
Does instruction improve outcomes?


---
class: middle, bg_butterfly

## Study design

--

`r emo::ji("raised_hand_with_fingers_splayed")` open-ended questions
 
--
 
`r emo::ji("v")` weeks to answer
  
--

`r emo::ji("point_up")` at the beginning of the semester,  `r emo::ji("point_up")` at the end
  

---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

.opaque[
Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?]

`r htmltools::withTags(br())`

--

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (individual)"),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Material:"), br(), br(), "Organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Mover:"), br(), br(), "DNA mutates, many organisms sexually reproduce, recombination", br(), br())))),
 tr(th("ULTIMATE (species)"),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Form:"), br(), br(), "Organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 0", br(), b("Formal:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Aristotle; Tinbergen, 1963; Nesse, 2013]]

---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions


Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`


```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (individual)"),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Material:"), br(), br(), "Organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Mover:"), br(), br(), "DNA mutates, many organisms sexually reproduce, recombination", br(), br())))),
 tr(th("ULTIMATE (species)"),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Form:"), br(), br(), "Organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 0", br(), b("Formal:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Aristotle; Tinbergen, 1963; Nesse, 2013]]


---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (individual)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Material:"), br(), br(), "Organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Mover:"), br(), br(), "DNA mutates, many organisms sexually reproduce, recombination", br(), br())))),
 tr(th("ULTIMATE (species)"),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Form:"), br(), br(), "Organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 0", br(), b("Formal:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Aristotle; Tinbergen, 1963; Nesse, 2013]]

---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (individual)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Material:"), br(), br(), "Organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Mover:"), br(), br(), "DNA mutates, many organisms sexually reproduce, recombination", br(), br())))),
 tr(th("ULTIMATE (species)"),
   td(htmltools::withTags(span(style = "opacity: 0", br(), b("Form:"), br(), br(), "Organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 0", br(), b("Formal:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Aristotle; Tinbergen, 1963; Nesse, 2013]]

---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (individual)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Material:"), br(), br(), "Organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Mover:"), br(), br(), "DNA mutates, many organisms sexually reproduce, recombination", br(), br())))),
 tr(th("ULTIMATE (species)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Form:"), br(), br(), "Organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 0", br(), b("Formal:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Aristotle; Tinbergen, 1963; Nesse, 2013]]


---
class: middle, left, bg_jelly

## Tinbergen`r htmltools::HTML("&rsquo;")`s Four Questions

Instead of having one species perfectly adapted to life on earth, there are millions of different species. Why is this the case?

`r htmltools::withTags(br())`

```{r, echo=FALSE, message = FALSE, warning = FALSE, eval = TRUE} 

htmltools::withTags(table(
 tr(th(htmltools::HTML("&nbsp;")),
   th("PATTERN (static)"),
   th("PROCESS (dynamic)")),
 tr(th("PROXIMATE (individual)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Material:"), br(), br(), "Organisms are constructed differently, have different DNA", br(), br()))),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Mover:"), br(), br(), "DNA mutates, many organisms sexually reproduce, recombination", br(), br())))),
 tr(th("ULTIMATE (species)"),
   td(htmltools::withTags(span(style = "opacity: 1", br(), b("Form:"), br(), br(), "Organisms take advantage of the environment in different ways", br(), br()))), 
 td(htmltools::withTags(span(style = "opacity: 1", br(), b("Formal:"), br(), br(), "Environmental heterogeneity and biological tradeoffs plus large timescales", br(), br()))))))

```

.footnote[
.small-font[Aristotle; Tinbergen, 1963; Nesse, 2013]]

---
class: middle, left

## What happened?

--

`r htmltools::withTags(img(src = "css/images/figs/cause-venn-diagram.png", width = "800px"))`

---
class: middle, left

## What happened?

.pull-left[
`r htmltools::withTags(img(src = "css/images/figs/pre-post-causes-empty.png", width = "450px"))`]

.pull-right-low[
`r htmltools::withTags(img(src = "css/images/figs/cause-venn-diagram.png", width = "390px"))`]

---
class: middle, left

## What happened?

.pull-left[
`r htmltools::withTags(img(src = "css/images/figs/pre-post-causes-r1.png", width = "450px"))`]

.pull-right-low[
`r htmltools::withTags(img(src = "css/images/figs/cause-venn-diagram.png", width = "390px"))`]

---
class: middle, left

## What happened?

.pull-left[
`r htmltools::withTags(img(src = "css/images/figs/pre-post-causes-r1-arrows.png", width = "450px"))`]

.pull-right-low[
`r htmltools::withTags(img(src = "css/images/figs/cause-venn-diagram.png", width = "390px"))`]

---
class: middle, left

## What happened?

.pull-left[
`r htmltools::withTags(img(src = "css/images/figs/pre-post-causes-final.png", width = "450px"))`]

.pull-right-low[
`r htmltools::withTags(img(src = "css/images/figs/cause-venn-diagram.png", width = "390px"))`]


---
class: top, left, bg_chaet

`r htmltools::withTags(br())`

## How are students constructing explanations?

--

.right-align[
.large-font[
.pink[
Schemas!]]]

.footnote[
Schank `r htmltools::withTags("&amp;")` Abelson, 1977; Chi _et al._, 2012]
---
class: top, left, bg_chaet

`r htmltools::withTags(br())`

### Reasoning by way of analogy to humans (anthropocentric):

`r htmltools::withTags(br())`

.footnote[
Coley `r htmltools::withTags("&amp;")` Tanner, 2015]
--


_There are so many different species of flowering plants because flowers are found all over the world just like humans, and they look different just like humans do._

`r htmltools::withTags(br())`
`r htmltools::withTags(br())`

.footnote[
Coley `r htmltools::withTags("&amp;")` Tanner, 2015]

---
class: top, left, bg_chaet

`r htmltools::withTags(br())`

### Reasoning by way of analogy to humans (anthropocentric):

`r htmltools::withTags(br())`

_Cave fish lose their eyesight because if you don`r htmltools::HTML("&rsquo;")`t use something the body will not maintain it, like if you stop exercising you lose your muscles._

`r htmltools::withTags(br())`
`r htmltools::withTags(br())`


.footnote[
Coley `r htmltools::withTags("&amp;")` Tanner, 2015]

---
class: top, left, bg_chaet

`r htmltools::withTags(br())`

### Reasoning by way of justifying the outcome (teleological):

`r htmltools::withTags(br())`

.footnote[
Coley `r htmltools::withTags("&amp;")` Tanner, 2015]
--

.italic[
There are so many different species on earth because if there were only one it would go extinct.]

`r htmltools::withTags(br())`
`r htmltools::withTags(br())`

.footnote[
Coley `r htmltools::withTags("&amp;")` Tanner, 2015]

---
class: top, left, bg_chaet

`r htmltools::withTags(br())`

### Reasoning by way of justifying the outcome (teleological):

`r htmltools::withTags(br())`

_Wood evolved because plants need it to survive in their environment._

`r htmltools::withTags(br())`
`r htmltools::withTags(br())`

.footnote[
Coley `r htmltools::withTags("&amp;")` Tanner, 2015]

--

.pull-right[
.large-font[
.pink[
Here, schemas are working against us!]]]

---
class: top, left, bg_2birds

## Using schemas to our advantage

`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`

.footnote[
Jolfaee _et al._, 2014]

--

```{r, echo=TRUE, warning=FALSE, error=FALSE, eval=TRUE}

coin.flip <- function(n){
   
   flips <- rbinom(n = n, size = 1, prob = 0.5)
   
   outcome <- ifelse(flips == 1, yes = "Heads", no = "Tails")
   
   table <- table(outcome) / n
   
   return(table)
   
}
```


.footnote[
Jolfaee _et al._, 2014]

---
class: top, left, bg_2birds

## Using schemas to our advantage

`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`


```{r, echo=TRUE, warning=FALSE, error=FALSE, eval=TRUE}
coin.flip(n = 8)
```

---
class: top, left, bg_2birds

## Using schemas to our advantage

`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`


```{r, echo=TRUE, warning=FALSE, error=FALSE, eval=TRUE}
coin.flip(n = 150)
```

---
class: top, left, bg_2birds

## Using schemas to our advantage

`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`
`r htmltools::withTags(br())`


```{r, echo=TRUE, warning=FALSE, error=FALSE, eval=TRUE}
coin.flip(n = 10000)
```

---
class: top, center, bg_gurnard

## Toward better biology education

--

.large-font[
`r htmltools::HTML('<i class="fas fa-map-marked-alt"></i>')`] 

.left-align[
Always bring facts and examples back to the bigger picture]

---
class: top, center, bg_gurnard

## Toward better biology education

.large-font[
`r htmltools::HTML('<i class="fas fa-map-marked-alt"></i>')`
`r htmltools::HTML("&nbsp;")`
`r htmltools::HTML('<i class="fas fa-code"></i>')`] 

.left-align[
Always bring facts and examples back to the bigger picture

Deal in concepts not in raw data]

---
class: top, center, bg_gurnard

## Toward better biology education

.large-font[
`r htmltools::HTML('<i class="fas fa-map-marked-alt"></i>')`
`r htmltools::HTML("&nbsp;")`
`r htmltools::HTML('<i class="fas fa-code"></i>')`
`r htmltools::HTML("&nbsp;")`
`r htmltools::HTML('<i class="fas fa-icons"></i>')`] 

.left-align[
Always bring facts and examples back to the bigger picture

Deal in concepts not in raw data

Use existing schemas to engage and challenge student thinking]

--
.bottom[
.large-font[
.pink[Thank you!]]]